Results

### 2016-02-16 doc/paper/results.Rmd
starting 2nd draft

---
title: "Results"
author: "JT McCrone"
date: "Febuary 9, 2016"
output: word_document
---
```{r.message=F,echo=F}
xamp_roc<-"figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"
table<-"table 1"
```
#### Accuracy

There are two major components to the accuracy of a variant caller, sensitivity and specificity. Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population. A sensitivity of 1 indicates that all true positives are found. Specificity, on the other hand, measures how well a variant caller can correctly identify true negatives. For example, the influenza genome is slightly larger than 13Kb. There are 3  potential variant nucleotides at each genomic position leaving a variant caller with over 40,000 potential variants that can be called. The vast majority of these potential variants will not be present in the starting population. They will only be found as the result of a sequencing or RT-PCR error. A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1. A perfect variant caller will have a sensitivity and specificity of 1 meaning it identifies all of the true variants and nothing else.

There is an obvious trade off between sensitivity and specificity. It may be the case that a variant caller has near perfect sensitivity, and  can find all the true positives in a sample. However, if in the process it has low specificity and identifies many false positives, downstream analysis will be confounded (`r table`). Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives it will have limited applications.

The trade off between sensitivity and specificity can be represented as a receiver operator curve or ROC (`r xamp_roc`). An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. To make a ROC for a variant caller one must test the assay on known data. The outcomes can then be stratified according to some metric that represents our belief that a given variant is real. Often this is a p value or quality score. Because the true variants are known, the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A). These points are then used to construct the curve (`r xamp_roc` B). A perfect ROC in which all the true positives can be separated from all potential false positives is a right angle that follows the upper left perimeter of the plot and passes through the point (0,1).

@@

There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a comprehensive comparison of their accuracy. Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy. In the process we hope to stress the importance of robustly benchmarking any variant caller under the experimental conditions to which it is being applied.

All variant callers attempt to distinguish true variants from errors by estimating  error rates across a given genome. These rates can take various metrics into account such as phred score, nucleotide position on a read, and the context of a given base call. The question then becomes is a putative variant found more often than would be expected given the applicable error rate. We have chosen DeepSNV as our variant caller of choice for this analysis. DeepSNV uses a clonal plasmid control to estimate an error rate for each possible nucleotide at each genomic position. The plasmid control contains is a clonal copy of the genome that is to be analyzed and is processed and sequenced in parallel with the test samples. Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process. DeepSNV has previously been benchmarked using a mixture of plasmids containing a 1.5Kb regions of the HIV pol gene. This process involved minimal PCR and was devoid of a reverse transcription step. Patient derived RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions. To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, were mixed at different frequencies to create known viral populations (`r miseq` A). Briefly, genomes were amplified separately in segment specific RT-PCR reactions. This was done to avoid biases in amplification that may skew frequency measurements towards one strain. Reconstituted cDNA genomes were made by mixing these products at equimolar ratios. The WSN33 mix was then serially diluted into PR8. These mixtures were sequenced using 2x250 reads on an Illumina Miseq and variants were called using DeepSNV.

Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline. `r Miseq` summarizes the accuracy of DeepSNV when applied to our experimental conditions. In the past DeepSNV was found to have a sensitivity of 86.0% and perfect specificity when applied to variants at 0.1%. Under our conditions  and with a slightly more stringent p value (p=0.01 versus p=0.05) we found a small decrease in both sensitivity and specificity when applied to variants at 0.63% (`r miseq` C). The  decrease in specificity is not surprising as our sequencing setup requires an error prone RT-PCR step. The drop in  sensitivity is likely due to the increased stringency expected when correcting for the number of test required to study a 13Kb region compared to the 1.5Kb region that was tested in the past. (_this can be tested_) It should also be noted that although the drop in specificity is small, 1 compared to 0.9983, the change in the number of false positives is fairly large (0 compared to 66).

#### Realistic conditions

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient derived samples are expected to be less divergent than the two strains used, and provide much less input than samples grown in cell culture. To mimic these conditions, 20 single point mutants were made in the WSN33 background and were diluted into wild type at decreasing frequencies (`r hiseq` A). This yielded 5 samples containing 20 known variants at frequencies ranging from 5% to 0.2% at a concentration of $10^5$ genomes/$\mu$l. These samples were diluted serially into viral media down to $10^3$ genomes/$\mu$l. This matches the range of inputs found in patient samples. The samples were then sequenced on an Illumina Hiseq and variants called using DeepSNV. Focusing on the $10^5$ genomes/$\mu$l samples we find that again sensitivity is correlated with variant frequency (`r hiseq` B and C). The fact that the sensitivity drops more dramatically than in our previous experiment is not surprising given the fact the the RNA input derived from cell culture samples is often orders of magnitude higher than that provided in this experiment. During library prep, a disproportionate amount of the 2.0% sample was lost compared to the others, and this loss likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0% sample. In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C). It should be noted that in all cases the specificity was greater than 99%. However, given the large number of potential false positives in an influenza genome (>40,000) even a specificity of 99% results in well over 200 erroneous variant calls. The drop in specificity is likely due to the decease in input levels. In our experience samples grown in cell culture have at least $10^7$ genomes/$\mu$l. As the genomic input decreases, we are more and more reliant on RT-PCR to amplify the true positive "signal" to a detectable level. It is not surprising that is error prone method also amplifies the noise in the process. Initially it is tempting to increase specificity by simply applying a more stringent p value cut off. This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B. However, it is apparent from  `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity. It is clear from `r hiseq` that under experimental conditions the DeepSNV output alone is insufficient to accurately separate true positives from false positives.

>>>>

---
title: "Results"
author: "JT McCrone"
date: "Febuary 15, 2016"
output: word_document
---
```{r.message=F,echo=F}
xamp_roc<-"figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"
table<-"table 1"
```

Accurately identifying rare single nucleotide variants depends on two components, sensitivity and specificity. Sensitivity, also known as the true positive rate, is the proportion of true variants that are found. Specificity, on the other hand, is the proportion of true negatives that are correctly identified. $1-\text{specificity}$ is the false positive rate, the ratio of false positive variants to true negatives. Both a sensitivity and specificity of 1 describe perfect accurarcy in which all the true variants are found and nothing else.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less strigent criteria in variant calling which reduces specificity. Conversly, increased stringency can improve specificity but reduce sensitivity. This relationship can be represented as a receiver operating characteristic (ROC) curve (`r xamp_roc`). An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. To make an ROC curve for a variant callling algorithm one must test the assay on known data. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score. In a controled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A). These points are then used to construct the curve (`r xamp_roc` B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

@@

It is beyond the scope of this work to comprehensively compare all variant calling methods virologists use. I Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy. In the process we demonstrate the importance of robustly benchmarking any variant calling pipeline under the experimental conditions to which it is being applied.

 We have chosen DeepSNV as our variant caller of choice for this analysis. DeepSNV uses a clonal, plasmid-derived, control to estimate local error rates controlling for sequence context accross the genome. Because the control is processed together with the experimental samples it has identical noise characteristics. DeepSNV applies a hierarchiacal binomial model at each genomic position and identifies true positive variants with frequencies greater than expected based on the control. Like many variant callers the accuracy of DeepSNV was initially determined using samples that required mimal PCR amplication. To characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus we creadted difed mixutres of two winfluenza strains, WSN33 and PR8 (`r miseq` A).

Because these viruses differ at 491 positions, there are 491 true positives at each dilution. On plasmids subserted to limited amplification, DeepSNV ad a sensitivity 0.860 and a specificity of 1 for SNV at 0.1% frequency. Under our realistic expermental conditions and with a slightly more stringent p value cutoff ( p=0.01 vs. p=0.05) we found a moderate reducttion in sensitivity  (0.830 for variants at 0.63% and 0.14 for variants at 0.16%). Our specificity was above 0.998 in all dilution. While this reduction in specificity appears minor, a false positive rate of 0.002 applied to the xx,xxx potential variants in the 13.xkb influenza genome results in 78 false positives.

#### An experimental intrahost population

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient-derived samples are expected to be less divergent than the two lab strains. Also the nucleic acid concentration in patient samples is much lower than that found samples grown in cell culture. To mimic these conditions, we generated 20 viral colones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform and cofirmed that no additional mutations >1% frequency had arisen between transfection and the passage 1 stock. We also quantified the genome copy number of each stock using quantitative PCR. We then mixed equal genome equivalents of these 20 virsuses, generating an artificial population of $10^5$ copies per microliter in which each mutation was present at 5% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations witha ll 20 mutations present a 2,1,0.5,0.2, and 0.1% frequency (`r hiseq` A). We then serially diluted each of these $10^5$ copy number populations into basal media to obtain populations with lower input. The range ($10^3-10^5$ copies per microliter) matches the inpus typically found in pateint-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate.

We considered the 20 mutations present in our initial viral mixture as the only true positives in our analysis. Any variant that was present in the wild type stock was masked and neither considered a true positive nor a true negative. This was done to avoid benchmarking DeepSNV with variants identified only be DeepSNV.

The samples with the highest input concentration had the best sensitivity (`r hiseq` B and C). Sensitivity dropped in the 0.5% frequency samples. In general, our sensitivity was lower than that seen previously in the cell culture samples where the input concentration was at least $10^7$ copies per microliter. The drop in sensitivity is likely due to the 100 fold decrease in variant genomes present in the $10^5$ samples. During library prep, a disproportionate amount of the 2.0% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0% sample.

In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C). It should be noted that in all cases the specificity was greater than 99%. However, given the large number of potential false positives in an influenza genome (>40,000) even a specificity of 99% results in well over 200 erroneous variant calls. The drop in specificity is likely due to the decease in input levels. In our experience samples grown in cell culture have at least $10^7$ genomes/$\mu$l. As the genomic input decreases, we are more and more reliant on RT-PCR to amplify the true positive "signal" to a detectable level. It is not surprising that is error prone method also amplifies the noise in the process. Initially it is tempting to increase specificity by simply applying a more stringent p value cut off. This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B. However, it is apparent from  `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity. It is clear from `r hiseq` that under experimental conditions the DeepSNV output alone is insufficient to accurately separate true positives from false positives.
